# Data Wrangling

Data wrangling refers to the way to view, change, and reorganize data in a data frame. Before we dive into the operations, it is important to know that there are two distinct schools of data wrangling: base R and `tidyverse`, where the latter has contributed substantially in enhancing the process. We will start with the base R version, followed by the `tidyverse`.

## Learning Objectives: {-}

This chapter covers common data wrangling techniques. After finishing this chapter, you should be able to:

+ Select rows that fulfill specified conditions.

+ Randomly sample rows from a data frame.

+ Sort rows in a data frame.

+ Select a subset of columns.

+ Add and remove columns to an existing data frame.

+ Tabulate data.

## Base R

### Select Rows by Condition(s)

The `subset()` function selects rows if they satisfy the condition(s). Here are the three relational operators to support multi-parts conditions:

1. "or" is denoted by `|`

2. "and" is denoted by `&`

3. "not" is denoted by `!`

Note that the operator for checking equality is "=="; two equal signs touching each other.

```{r}
# Choose all setosa iris. 
setosa_iris <- subset(iris, Species == 'setosa')

# How many are selected?
print(paste("There are", nrow(setosa_iris), "rows."))
```

```{r}
# Peek into the first few rows of the result.
head(setosa_iris)
```

```{r}
# More than one condition. The relational operator is the logical OR, i.e., a vertical bar |
iris_or_eg <- subset(iris, Sepal.Length > 5 | Sepal.Width < 3 )

# How many?
print(paste("There are", nrow(iris_or_eg), "rows."))

head(iris_or_eg)
```

```{r}
# More than one condition. The relational AND operator.
iris_and_eg <- subset(iris, Petal.Length > 6 & Petal.Width < 2 )

# How many?
print(paste("There are", nrow(iris_and_eg), "rows."))

iris_and_eg
```

```{r}
# No setosa iris
iris_no_setosa_eg <- subset(iris, Species != 'setosa')

# How many?
print(paste("There are", nrow(iris_no_setosa_eg), "rows."))

head(iris_no_setosa_eg)
```

### Select Rows Randomly

It is quite common to randomly sample a subset of rows from a data frame. It takes to two steps to achieve that:

**Step 1.** Create a list (vector) of random row indices by the `sample()` function. Specify the following parameters: 

a.) The range of values. For this example, the range is from 1 to the last row of `iris`, which is the return of `nrow(iris)`.

b) How many samples, e.g., `size = 5`.

c) Can a sample be drawn repeatedly? If not, `replace = F`.

```{r}
# seq(nrow(iris)) generates 1, 2, 3, ..., 150.
rnd_idx <- sample(seq(nrow(iris)), size = 5, replace = F)
rnd_idx
```

Step 2: select rows from iris according to the indices in `rnd_idx`.

```{r}
a_random_sample <- iris[rnd_idx,]

a_random_sample
```

### Sort Rows

It is a two-step process similar to the random sampling of rows from the previus subsection. Suppose you want to sort `iris` in `Sepal.Length` in ascending order (small to large).

Step 1: Obtain a list of row indices that indicates the sorting order to the data.

```{r}
sorted_idx <- order(iris$Sepal.Length)
sorted_idx
```

Step 2: Rearrange the rows in the data frame according to the sorted row indices.

```{r}
# The 5 shortest Sepal Lengths
head(iris[sorted_idx,])

# The 5 longest Sepal Lengths
tail(iris[sorted_idx,])
```
Note that the above two steps do NOT change the original order of the rows in the data frame. If you intend to change the row order permanently, use the assignment operator to update the existing data frame or create a new data frame to store the sorted rows. See below:

```{r}
ascending_iris <- iris[sorted_idx,]

# The shortest Sepal Length
head(ascending_iris)

# The longest Sepal Length
tail(ascending_iris)
```

To sort rows in descending order (large to small), add the parameter `decreasing = T` to the `order()` function.

```{r}
desc_idx <- order(iris$Sepal.Length, decreasing = T)
desc_idx

decreasing_iris <- iris[desc_idx,]
head(decreasing_iris)
tail(decreasing_iris)
```

### Rearranging Columns

Suppose you want to move the `Species` column to the leftmost position. It can be done using two approaches.

```{r}
# Reorder columns by name
new_iris <- iris[,c("Species","Sepal.Length","Sepal.Width","Petal.Length","Petal.Width")]
head(new_iris)
```

```{r}
# Reorder columns by index
new_iris <- iris[,c(5,1,2,3,4)]
head(new_iris)
```

Note that if a data frame contains many columns, say >100, it is daunting to use this approach. A much better way can be done using `dplyr`, so stay tuned.

### Focus on a Subset of Columns

If a data frame contains many columns, way more than you need for analysis. It is neat to focus on those columns and remove the rest. Let's suppose you want to focus only on sepal information and species.

```{r}
sepal_iris <- iris[,c("Sepal.Length","Sepal.Width","Species")]
dim(sepal_iris)

head(sepal_iris)
```

### Add a New Column

Suppose we want to add a new column called `Sepal.Ratio`, which is the ratio `Sepal.Length` to `Sepal.Width`. Here is the code:

```{r}
iris$Sepal.Ratio <- iris$Sepal.Length/iris$Sepal.Width
head(iris)
```

You can see the new column `Sepal.Ratio` is added to the rightmost position.

### Remove a Column

You can remove a column from a data frame, e.g., remove the `Sepal.Ratio` added above.

```{r}
iris$Sepal.Ratio <- NULL

# the column is gone
head(iris)
```

### Tabulation {#base-tabulation}

Tally the number of samples per one or more categorical variables (factors) by the `table()` function.

```{r}
table(iris$Species)

# Alternatively, use with()
with(iris, table(Species))
```
It can tally more than one categorical variable. Let's us another built-in dataset `CO2` to illustrate the point.

```{r}
with(CO2, table(Plant, Type))
```

## `tidyverse` 

There is an alternate, and importantly, less cryptic way to perform data wrangling. This new set of functions are offered by the R package `dplyr`. One design goal of `dplyr` is to use **verbs** to name these operations, a huge difference from the unpronounceable symbols, such as `$`, `[`, etc. by base R as shown in the previous section.

Before we dive into `dplyr`, it is noteworthy to mention that there is a slew of packages with the aim of improving data cleaning, data visualization, and loading data in various formats. To spare people from the question of when to use what package, all these packages are bundled in a single package called `tidyverse`. Once you have activated `tidyverse`, `dplyr`, `ggplot` (for plotting data), and other packages are available for use. When you call a function, you don't have to worry if it is from `dplyr` or `tidyr` because the `tidyverse` community has resolved all name conflicts. An analogy is like there is only one "Main Street" in `tidyverse` package; no ambiguity at all.  **The take home is: just activate `tidyverse` and you are good to go.**

`tidyverse` does not come with the base R so you have to install it manually. But the installation is simple, click the menu option Tools, select the first option "install packages", type "tidyverse", and click the "Install" button.

Let's activate `tidyverse` by using the `library()` function. Alternatively, you can use `require()`; they are largely doing the same thing. Note that you only need to activate a package once per session.

```{r}
library(tidyverse)
```

The `library()` function has churned out many messages telling you what packages (9 of them) are loaded together. 

Let's use `tidyverse` to repeat the data wrangling tasks by base R above.

### Select Rows by Condition(s)

The function is `filter()`. It chooses rows that satisfy the specified condition(s). The syntax is `filter(data_frame, conditions)`. E.g., 

```{r}
# select all setosa iris
setosa_iris <- filter(iris, Species == 'setosa')

# How many?
print(paste("There are", nrow(setosa_iris), "rows."))
```

```{r}
# More than one condition. The symbol for the logical OR operator is a vertical bar |
iris_or_eg <- filter(iris, Sepal.Length > 5 | Sepal.Width < 3 )

# How many?
print(paste("There are", nrow(iris_or_eg), "rows."))

head(iris_or_eg)
```

```{r}
# More than one condition. The symbol for the logical AND operator is &
iris_and_eg <- filter(iris, Petal.Length > 6 & Petal.Width < 2 )

# How many?
print(paste("There are", nrow(iris_and_eg), "rows."))

iris_and_eg
```

```{r}
# No setosa iris
iris_not_eg <- filter(iris, Species != 'setosa')

# How many?
print(paste("There are", nrow(iris_not_eg), "rows."))

head(iris_not_eg)
```

### Select Rows Randomly

With `tidyverse`, you can do it in one step by using the `sample_n()` function.

```{r}
a_random_sample <- sample_n(iris, size = 5)

a_random_sample
```

### Sort Rows in a Data Frame

Use `arrange()`. Note that the function will not change the original data frame. Use the assignment operator (`<-`) to change the original data frame or store the changes in a new data frame. For example, sort `iris` by `Sepal.Length`, and store the sorted result in a new data frame.

```{r}
sorted_iris <- arrange(iris, Sepal.Length)

# The shortest Sepal Length
head(sorted_iris)

# The longest Sepal Length
tail(sorted_iris)
```

To sort rows according to the descending order of the sorting column, simply mark the sorting column with a minus `-`. E.g.,

```{r}
decreasing_iris <- arrange(iris, -Sepal.Length)
head(decreasing_iris)
tail(decreasing_iris)
```

Specify more than one sorting column is much easier in `dplyr` than base R. E.g. sort `iris` in descending order `Sepal.Length` and ascending order of `Petal.Length`.

```{r}
double_sorted_iris <- arrange(iris, -Sepal.Length, Petal.Length)
head(double_sorted_iris)
tail(double_sorted_iris)
```

For rows with equal `Sepal.Length` (rows 2-5, 147-149), you can see their `Petal.Length` are in ascending order.

### Rearranging columns of a data frame

Suppose you want to move the `Species` column to the leftmost position. You're going to love `dplyr`, as it can be done easily by using two functions: `select()` and `everything()`.

```{r}
# Reorder columns by name
new_iris <- select(iris, Species, everything())
head(new_iris)
```

### Focus on a Subset of Columns

If a data frame contains many columns, way more than you need for analysis. It is neat to focus on those columns and remove the rest. E.g., use the `select()` function to keep only sepal information and species.

```{r}
sepal_iris <- select(iris, Sepal.Length, Sepal.Width, Species)
dim(sepal_iris)

head(sepal_iris)
```

### Add a New Column

Use the `mutate()` function add columns to a data frame.

```{r}
iris <- mutate(iris, Sepal.Ratio = Sepal.Length/Sepal.Width)
head(iris)
```

### Remove a Column

Use the `select()` function as before but put a minus `'-'` symbol before the column name to be removed.

```{r}
iris <- select(iris, -Sepal.Ratio)
head(iris)
```

### Rename a Column

So far, you have seen all the data wrangling tasks performed by base R using `tidyverse`. Starting from here, you are going to demonstrate tasks that are harder to do with base R. The first task is to rename a column. Suppose you want to rename `Species` to `Iris_Species`. You can do it by the `rename()` function:

```{r}
# new_name = existing_name
iris <- rename(iris, Iris_Species = Species)
head(iris)
```

Let's revert the renaming so it will not affect the following examples.

```{r}
iris <- rename(iris, Species = Iris_Species)
head(iris)
```

### Pipe

Often, data wrangling is a multi-step process in which many intermediate steps are involved before producing the final result. E.g., select iris where its `Petal.Length` is between 0.5 and 2, and sort the results in descending order of `Petal.Length`. As you can see, it involves filtering, followed by sorting. It can done into two steps via an intermediate data frame as below:

```{r}
tmp_iris <- filter(iris, Petal.Length >= 0.5 & Petal.Length <=2)
final_iris <- arrange(tmp_iris, -Petal.Length)
dim(final_iris)
head(final_iris)
```

The final result contains 50 rows.

`tidyverse` provides a neater way - no intermediate data frames - to do it by the concept of **pip**. You can imagine it like the assembly line of a factory, where an upstream work station passes (pipe) intermediate results to the next work station for further processing.

The pip symbol is represented by **`%>%`**, no space between them. Let's see how to apply pip to the above example without creating the intermediate data frame `tmp_iris`.

```{r}
final_results <- iris %>% 
                   filter(Petal.Length >= 0.5 & Petal.Length <=2) %>%
                   arrange(-Petal.Length)
dim(final_iris)
head(final_iris)
```

Let's walk through the code above line by line.

+ `iris %>%` sends all rows, 150 in total, to the downstream `filter()` function.

+ `filter(Petal.Length >= 0.5 & Petal.Length <=2) %>%` receives the data, and check each row if it satisfies the specified conditions. If so, send it to the downstream process; otherwise do not pass down.

+ `arrange(-Petal.Length)` received rows that satisfy the previous filtering process, and sort them in descending order of `Petal.Length`.

+ `final_results <-` the sorted result is assigned to the variable `final_results`.

By using pip (`%>%`), you can avoid intermediate variables, making the code cleaner and easily to read.

### Tabulation and Summarize {#tidyverse-tabulation}

Another task which is useful but difficult to get it done in base R is to produce summary information by group. E.g., you want to produce a table that shows the average sepal length for each species. Here's how to do it using two new functions `group_by()` and `summarize()`, and pipe them together.

E.g., tally the number of samples by `Plant` and 'Type' in `CO2`. The function that does the tallying is `n()`.

```{r}
CO2 %>%
  group_by(Plant, Type) %>%
  summarize(n=n())
```

You do not see much difference between base R and `tidyverse` in the example above. But if you want to do more than tallying, `summarize()` is a treat, as `summarize()` supports the descriptive statistics functions mentioned before, such as mean(), sd(), median(), quantile(), etc. E.g,

```{r}
iris %>%
  group_by(Species) %>%
  summarize(u=mean(Sepal.Length))
```

You can apply multiple functions in `summarize()`. E.g., shows the number of samples and standard deviation.

```{r}
iris %>%
  group_by(Species) %>%
  summarize(n=n(),
            u=mean(Sepal.Length),
            s=sd(Sepal.Length))
```

You can group by more than one column. Let's switch to `CO2` dataset to illustrate this point.

```{r}
CO2 %>%
  group_by(Plant, Type, Treatment) %>%
  summarize(n=n(),
            u=mean(conc),
            s=sd(conc))
```

**Two remarks:**

1. `group_by()` is never used alone. It is often followed by `summarize()`.

2. The variables feed to `group_by()` must be factor, i.e., categorical.


### A Summary of `tidyverse` functions

- `filter()`: select rows that satistify the specified conditions.

- `select()`: choose or drop certain columns.

- `arrange()`: sort rows by the specified columns.

- `rename()`: change the name of a column.

- `mutate()`: add a column or modify the values of an existing column.

- `group_by()`, followed by `summarize()`

- pip `%>%`: funnel intermediate result from an upstream process to the input of an immediate downstream process.

## Summary

In this chapter, you have seen how to perform data wrangling with basic R and `tidyverse`. What you have seen is only a fraction of what `tidyverse` can do. Readers who want to explore the full capability of `tidyverse` should consult the most authorative source: **R for Data Science** by Hadley Wickham & Garrett Grolemund (https://r4ds.had.co.nz/). In the next chapter, you will see how to plot data. A picture is worth a thousand words. Data visualization is an indispensible step for data analysis. But before the next topic, let's recall all the new R commands discussed in this chapter.

<!--
`|`, `&`, `==`, `!`, `!=`, `subset()`, `sample()`, `order()`, `NONE` (a keyword), `table()`, `library()`, `require()`, `filter()`, `select()`, `arrange()`, `rename()`, `mutate()`, `group_by()`, `summarize()`, `n()`, and pip `%>%`.
-->

```{r echo=FALSE, warning=FALSE}
cmds <- c("|", "`&`", "`==`", "`!`", "`!=`", "`subset()`", "`sample()`", "`order()`", "`NONE` (a keyword)", "`table()`", "`library()`", "`require()`", "`filter()`", "`select()`", "`arrange()`", "`rename()`", "`mutate()`", "`group_by()`", "`summarize()`", "`n()`","pip `%>%`", "", "", "") # fill up the matrix with empty strings

cmds_mat <- matrix(cmds, ncol = 4)
knitr::kable(
  cmds_mat, booktabs = TRUE,
  caption = "A Summary of R Operators and Functions"
)
```