# Data Wrangling {#sec-02-data-wrangling}

Data wrangling refers to the process of viewing, changing, and reorganizing rows and columns in a data frame. Before we dive into the operations, it is important to know that there are two distinct schools of data wrangling: base R and `tidyverse`. The base R approach means that no additional packages are required, but the `tidyverse` approach requires the installation of `tidyverse`, even though the installation is very straightforward. In my opinion, `tidyverse` has substantially refined and enhanced data wrangling. You will see it in this chapter. Regardless, we will discuss both approaches. Let's start with the base R version.

## Learning Objectives: {.unnumbered}

This chapter covers common data wrangling techniques. After finishing this chapter, you should be able to:

- Select rows that fulfill specified conditions.

- Randomly sample rows from a data frame.

- Sort rows in a data frame.

- Select a subset of columns.

- Add and remove columns to an existing data frame.

- Tabulate data.

## Base R

### Select Rows by Condition(s)

The `subset()` function selects rows if they satisfy the condition(s). Below are six relational operators:

1.  `>` greater than.

2.  `>=`[^02_data_wrangling-1] greater than or equal to.

3.  `<` less than.

4.  `<=`[^02_data_wrangling-2] less than or equal to.

5.  `==`[^02_data_wrangling-3] equal to.

6.  `!=`[^02_data_wrangling-4] not equal to.

[^02_data_wrangling-1]: No spaces between the two symbols.

[^02_data_wrangling-2]: No spaces between the two symbols.

[^02_data_wrangling-3]: No spaces between the two symbols.

[^02_data_wrangling-4]: No spaces between the two symbols.

If the selection involves multiple conditions, they must be joined by relational operators. Here are the three logical operators required for multi-part conditions:

1.  `|` denotes the logical OR.

2.  `&` denotes the logical AND.

3.  `!` denotes the logical NOT.

#### `subset()` with Vectors

Let's define a vector.

```{r}
# A sample of pumpkins' weights
pumpkins <- c(6.85,6.56,6.23,7.55,4.45,10.14,3.52,6.62,3.76,7.62, 7)
```

Select pumpkins heavier than 7 lbs.

```{r}
subset(pumpkins, pumpkins>7)
```

Select pumpkins that are exactly 7 lbs.

```{r}
subset(pumpkins, pumpkins==7)
```

Select pumpkins between 6 and 7 lbs, inclusive.

```{r}
subset(pumpkins, pumpkins>=6 & pumpkins<=7)
```

#### `subset()` with Data Frames

Here is an example that chooses only the `setosa` species from `iris` data frame and keeps them in a new data frame. Note that the equality check is case-sensitive.

```{r}
setosa_iris <- subset(iris, Species == 'setosa')

# How many are selected?
print(paste("There are", nrow(setosa_iris), "rows."))
```

```{r}
# Peek into the first few rows of the result.
head(setosa_iris)
```

More than one condition. Select rows with `Sepal.Length` is greater than 5 or `Sepal.Width` is smaller than 3.

```{r}
iris_or_eg <- subset(iris, Sepal.Length > 5 | Sepal.Width < 3 )

# How many?
print(paste("There are", nrow(iris_or_eg), "rows."))

head(iris_or_eg)
```

Select rows with `Petal.Length` is greater than 6 and `Petal.Width` is smaller than 2.

```{r}
# More than one condition. The relational AND operator.
iris_and_eg <- subset(iris, Petal.Length > 6 & Petal.Width < 2 )

# How many?
print(paste("There are", nrow(iris_and_eg), "rows."))

iris_and_eg
```

Select everything except for `setosa`.

```{r}
# No setosa iris
iris_no_setosa_eg <- subset(iris, Species != 'setosa')

# How many?
print(paste("There are", nrow(iris_no_setosa_eg), "rows."))

head(iris_no_setosa_eg)
```

### Random Sampling

It is quite common to randomly sample a subset of items from a vector or rows from a data frame.

#### Sample Items Randomly from a Vector

```{r}
# A sample of pumpkins' weights
pumpkins <- c(6.85,6.56,6.23,7.55,4.45,10.14,3.52,6.62,3.76,7.62, 7)
```

Randomly pick 5 samples without replacement.

```{r}
sample(pumpkins, size=5, replace=F)
```

The `replace=` option can be twisted to simulate tossing a coin N times. E.g.,

```{r}
sample(c('H','T'), size=10, replace=T)
```

#### Sample Rows Randomly from a Data Frame

Let's say we want to pick 5 samples randomly from `iris`. It takes two steps to do that:

**Step 1:** Create a list (vector) of random row indexes by the `sample()` function. Specify the following parameters:

a.) The range of indexes. In this example, the range is from 1 to the last row of `iris`, which can be obtained from `nrow(iris)`.

b)  How many samples, e.g., `size = 5`.

c)  Can a sample be drawn repeatedly? If not, `replace = F`.

```{r}
rnd_idx <- sample(seq(nrow(iris)), size = 5, replace = F)
rnd_idx
```

`seq(nrow(iris))` generates a vector containing 1, 2, 3, ..., 150.

**Step 2:** Select rows from iris according to the indexes in `rnd_idx`.

```{r}
a_random_sample <- iris[rnd_idx,]

a_random_sample
```

The indexing method is based on the previous chapter. Check out here: @sec-access-rows-columns.

### Sort Items

#### Sort Items in a Vector

It can be done using the `sort()` function.

```{r}
sort(pumpkins)
```

Sort it in descending order (large to small).

```{r}
sort(pumpkins, decreasing = TRUE)
```

#### Sort Rows in a Data Frame

Unfortunately, the `sort()` function does not work for the two-dimensional data frame. We have to use a method similar to sampling rows. It takes two steps to sort rows in a certain order. Suppose you want to sort rows in `iris` by `Sepal.Length` in ascending order (small to large).

**Step 1:** Obtain a list (vector) of row indexes that reflects the sorting order of the data frame.

```{r}
sorted_idx <- order(iris$Sepal.Length)
sorted_idx
```

**Step 2:** Rearrange the rows in the data frame according to the sorted row indexes.

```{r}
# The 5 shortest Sepal Lengths
head(iris[sorted_idx,])

# The 5 longest Sepal Lengths
tail(iris[sorted_idx,])
```

Note that the steps above do NOT change the original order of the rows in the data frame. If you intend to change the row order permanently, use the assignment operator to update the existing data frame or create a new data frame to store the sorted rows. See below:

```{r}
ascending_iris <- iris[sorted_idx,]

# The shortest Sepal Length
head(ascending_iris)

# The longest Sepal Length
tail(ascending_iris)
```

To sort rows in descending order (large to small), add the parameter `decreasing = T` to the `order()` function.

```{r}
desc_idx <- order(iris$Sepal.Length, decreasing = T)
desc_idx

decreasing_iris <- iris[desc_idx,]
head(decreasing_iris)
tail(decreasing_iris)
```

### Rearranging Columns

Suppose you want to move the `Species` column to the leftmost position. It can be done in two ways.

```{r}
# Reorder columns by name
new_iris <- iris[,c("Species","Sepal.Length","Sepal.Width","Petal.Length","Petal.Width")]
head(new_iris)
```

```{r}
# Reorder columns by index
new_iris <- iris[,c(5,1,2,3,4)]
head(new_iris)
```

Note that if a data frame contains many columns, say \>100, this method is daunting. A much better way can be done using `tidyverse`, so stay tuned.

### Focus on a Subset of Columns

Suppose a data frame contains many columns, way more than you need for analysis. It is neat to focus on those columns and leave out the rest. Let's suppose you want to focus only on sepal information and species.

```{r}
sepal_iris <- iris[,c("Sepal.Length","Sepal.Width","Species")]
dim(sepal_iris)

head(sepal_iris)
```

### Add a New Column

Suppose we want to add a new column called `Sepal.Ratio`, which is the ratio `Sepal.Length` to `Sepal.Width`. Here is the code:

```{r}
iris$Sepal.Ratio <- iris$Sepal.Length/iris$Sepal.Width
head(iris)
```

You can see the new column `Sepal.Ratio` is added to the rightmost position.

Suppose you want to add a new column that records the name of the researcher who collected the sample. If one researcher does it, you don't have to repeat the same name in a vector. Here is the code:

```{r}
iris$Researcher <- "Brian"
head(iris)
```

### Remove a Column

You can remove a column from a data frame, e.g., remove the `Sepal.Ratio` and `Researcher` added above.

```{r}
iris$Sepal.Ratio <- NULL
iris$Researcher <- NULL

# the columns are gone
head(iris)
```

### Tabulation {#sec-base-tabulation}

Tally the number of samples per one or more categorical variables (factors) by the `table()` function.

```{r}
table(iris$Species)

# Alternatively, use with()
with(iris, table(Species))
```

The data frame can be tallied by more than one categorical variable. Let's use another built-in dataset `CO2` to illustrate the point.

```{r}
with(CO2, table(Plant, Type))
```

That's all for data wrangling using base R.

## Exercises for Base R

Q1. Use the built-in data set `islands` to answer this question. `islands` is a named vector containing the landmass areas of the world's major continents and islands, where the unit of landmass is in thousands of square miles.

a)  Show places with landmass less than 50.

b)  Show places with landmass between 100 and 200, inclusive.

c)  Show places with landmass either greater than 1000 or smaller than 100, inclusive.

d)  Show places with landmass 30.

e)  Randomly sample 10 places with no replacement.

f)  Show the five largest places.

g)  Show the five smallest places.

```{r echo=F, eval=F}
#a
print("Q1a")
subset(islands, islands < 50)

#b
print("Q1b")
subset(islands, islands >= 100 & islands <= 200)

#c
print("Q1c")
subset(islands, islands >= 5000 | islands <= 50)

#d
print("Q1d")
subset(islands, islands == 30)

#e
print("Q1e")
sample(islands, size = 10, replace = F)

#f
print("Q1f")
idx <- order(islands, decreasing = T)
islands[idx][1:5]

#g
print("Q1g")
idx <- order(islands)
islands[idx][1:5]
```

Q2. Use the built-in data set `mtcar` to answer this question.

a.  Show cars with more than 4 gears `gear`.

b.  Show cars with `mpg` between 15 and 25, inclusive.

c.  Show cars with horsepower `hp` greater than 100 and the number of cylinders `cyl` less than 6.

d.  Show cars that do not have 6 cylinders.

e.  `qsec` stands for quarter-mile time in seconds. It measures acceleration; the shorter the time, the faster the acceleration. Show three cars that have the highest acceleration.

f.  Show three cars that have the lowest acceleration.

g.  Show two cars whose accelerations are closest to the median.

```{r echo=F, eval=F}
#a
subset(mtcars, gear>4)

#b
subset(mtcars, mpg>=15 & mpg<=25)

#c
subset(mtcars, hp>100 & cyl<6)

#d
subset(mtcars, cyl != 6)

#e
idx <- order(mtcars$qsec)
mtcars[idx,][1:3,]

#f
idx <- order(mtcars$qsec)
mtcars[idx,][1:3,]

#g
median_qsec <- median(mtcars$qsec)
idx <- order(mtcars$qsec)
# 32 cars, 16th and 17th cars are ranked in the middle
mtcars[idx,][16:17,]
```

Q3. Rearranging columns.

a.  Create a new data frame called `mtcars_small` that contains only three columns: `mpg`, `cyl`, and `wt`. FYI, `wt` is weight in 1,000 lbs.

b.  Create a new column in `mtcars_small` called `mpg_by_wt`, which is defined as `mpg` divided by `wt`.

c.  Show the top three cars with the highest `mpg` by `wt`.

d.  Remove `mpg_by_wt` from `mtcars_small`.

e.  Create a new variable in `mtcars_small` called `efficiency`. If the `mpg` of a car is greater than 20, set `efficiency` to `"gas guzzler"`; otherwise, set it to `"economy"`. Tabulate `efficiency`.

```{r echo=F, eval=F}
#a
mtcars_small <- mtcars[,c('mpg','cyl','wt')]
head(mtcars_small)

#b
mtcars_small$mpg_by_wt <- mtcars_small$mpg/mtcars_small$wt
head(mtcars_small)
# Below is a common mistake. 
# mpg_by_wt <- mtcars_small$mpg/mtcars_small$wt
#
# It is wrong because the variable mpg_by_wt was created out of
# mtcars_small. When you look at mtcars_small, mpg_by_wt is not there.

#c
indx <- order(mtcars_small$mpg_by_wt, decreasing = T)
mtcars_small[idx,][1:3,]

#d
mtcars_small$mpg_by_wt <- NULL
head(mtcars_small)

#e
mtcars$efficiency <- ifelse(mtcars$mpg > 20, "economy", "gas guzzler")

# 2. Convert the new character column into a factor
mtcars$efficiency <- factor(mtcars$efficiency)

# View the result
table(mtcars$efficiency)

```

## `tidyverse`

`tidyverse` is an alternative, and importantly, less cryptic way to perform data wrangling on data frames. One design goal of is to name functions using **verbs**, which improves code readability, setting it apart from using unpronounceable symbols, such as `$`, `[`, etc., as used by base R as shown in the previous section. **However, `tidyverse` works with data frames only.** You still need base R to process vectors.

Before we dive into `tidyverse`, it is noteworthy to clarify that `tidyverse` is an integrated ecosystem of R packages designed for data science. Two packages actually provide the data wrangling functions discussed below: `dplyr`, and `tidyr`. As this book is a little book, you can assume the umbrella package provides everything.

As `tidyverse` is an add-on package, you have to install it manually. But the installation is simple. In RStudio, click the *Tools* menu option, select the first option: *Install Packages*. In the *Packages* box, type "tidyverse", and click the "Install" button. You only need to install it once, until after a major R upgrade.

### Activate `tidyverse`

After opening a new session, you need to activate `tidyverse` using the `require()` function.

```{r}
require(tidyverse)
```

The `require()` function prints messages telling you what packages (9 of them) are loaded together.

Alternatively, you can use the `library()` function to activate a package. Putting aside technical nuances, these two functions are largely doing the same thing. Since activation is an action, using a verb (`require()`) makes more sense than using a noun (`library()`). Note that you only need to activate a package once per R session.

Let's use `tidyverse` to repeat the data wrangling tasks in base R above.

### Select Rows by Condition(s)

The function is `filter()`. It chooses rows that satisfy the specified condition(s). The syntax is `filter(data_frame, conditions)`. E.g.,

```{r}
# select all setosa iris
setosa_iris <- filter(iris, Species == 'setosa')

# How many?
print(paste("There are", nrow(setosa_iris), "rows."))
```

```{r}
# More than one condition. The symbol for the logical OR operator is a vertical bar |
iris_or_eg <- filter(iris, Sepal.Length > 5 | Sepal.Width < 3 )

# How many?
print(paste("There are", nrow(iris_or_eg), "rows."))

head(iris_or_eg)
```

Specifying more than one condition. Use a relational operator to join conditions together.

```{r}
iris_and_eg <- filter(iris, Petal.Length > 6 & Petal.Width < 2 )

# How many?
print(paste("There are", nrow(iris_and_eg), "rows."))

iris_and_eg
```

```{r}
# No setosa iris
iris_not_eg <- filter(iris, Species != 'setosa')

# How many?
print(paste("There are", nrow(iris_not_eg), "rows."))

head(iris_not_eg)
```

### Select Rows Randomly

With `tidyverse`, you can do it in one step by using the `sample_n()` function.

```{r}
a_random_sample <- sample_n(iris, size = 5)

a_random_sample
```

### Sort Rows in a Data Frame

Use `arrange()`. Note that the function will not change the original data frame. Use the assignment operator (`<-`) to make permanent changes to the original data frame or store the changes in a new data frame. For example, sort `iris` by `Sepal.Length`, and store the sorted result in a new data frame.

```{r}
sorted_iris <- arrange(iris, Sepal.Length)

# The shortest Sepal Length
head(sorted_iris)

# The longest Sepal Length
tail(sorted_iris)
```

To sort rows in descending order, mark the sorting column with a minus `-`. E.g.,

```{r}
decreasing_iris <- arrange(iris, -Sepal.Length)
head(decreasing_iris)
tail(decreasing_iris)
```

Specifying more than one sorting column is much easier in `tidyverse` than base R. E.g. sort `iris` in descending order of `Sepal.Length` and ascending order of `Petal.Length`.

```{r}
double_sorted_iris <- arrange(iris, -Sepal.Length, Petal.Length)
head(double_sorted_iris)
tail(double_sorted_iris)
```

For rows with equal `Sepal.Length` (rows 2-5, 147-149), you can see their `Petal.Length` are in ascending order.

### Reorganize columns of a data frame

Suppose you want to move the `Species` column to the leftmost position. You're going to love `tidyverse`, as it can be done easily by using two functions: `select()` and `everything()`.

```{r}
# Reorder columns by name
new_iris <- select(iris, Species, everything())
head(new_iris)
```

### Focus on a Subset of Columns

If a data frame contains many columns, way more than you need for analysis. It is neat to focus on those columns and remove the rest. E.g., use the `select()` function to keep only sepal information and species.

```{r}
sepal_iris <- select(iris, Sepal.Length, Sepal.Width, Species)
dim(sepal_iris)

head(sepal_iris)
```

### Add a New Column

Use the `mutate()` function to add columns to a data frame.

```{r}
iris <- mutate(iris, Sepal.Ratio = Sepal.Length/Sepal.Width)
head(iris)
```

### Remove a Column

Use the `select()` function and put a minus `'-'` symbol before the column name(s) to be removed.

```{r}
iris <- select(iris, -Sepal.Ratio)
head(iris)
```

### Rename a Column

So far, you have seen all the data wrangling tasks performed by base R using `tidyverse`. Starting from here, you are going to demonstrate tasks that are harder to do with base R. The first task is to rename a column. Suppose you want to rename `Species` to `Iris_Species`. You can do it with the `rename()` function:

```{r}
# new_name = existing_name
iris <- rename(iris, Iris_Species = Species)
head(iris)
```

Let's revert the renaming so it will not affect the following examples.

```{r}
iris <- rename(iris, Species = Iris_Species)
head(iris)
```

### Pipe

Often, data wrangling is a multi-step process with many intermediate steps before producing the final result. E.g., select iris where its `Petal.Length` is between 0.5 and 2, and sort the results in descending order of `Petal.Length`. As you can see, it involves filtering, followed by sorting. It can be done in two steps via an intermediate data frame as below:

```{r}
tmp_iris <- filter(iris, Petal.Length >= 0.5 & Petal.Length <=2)
final_iris <- arrange(tmp_iris, -Petal.Length)
dim(final_iris)
head(final_iris)
```

The final result contains 50 rows.

`tidyverse` provides a neater way with no intermediate data frames produced. What is new is the **pipe** concept. You can imagine a pipe is like the assembly line of a factory, where an upstream workstation passes (pipes) intermediate results to the next workstation for further processing.

The pipe symbol is represented by **`%>%`**, with no space between them. Let's see how to apply the pipe to the above example without creating the intermediate data frame `tmp_iris`.

```{r}
final_results <- iris %>% 
                   filter(Petal.Length >= 0.5 & Petal.Length <=2) %>%
                   arrange(-Petal.Length)
dim(final_iris)
head(final_iris)
```

Let's walk through the code above line by line.

- `iris %>%` sends all rows, 150 in total, to the downstream `filter()` function.

- `filter(Petal.Length >= 0.5 & Petal.Length <=2) %>%` receives the data and checks each row to see if it satisfies the specified conditions. If so, it sends it to the downstream process; otherwise, it does not pass it down.

- `arrange(-Petal.Length)` received rows that satisfy the previous filtering process, and sort them in descending order of `Petal.Length`.

- `final_results <-` the sorted result is assigned to the variable `final_results`.

By using pipe (`%>%`), you can avoid intermediate variables, making the code cleaner and easier to read.

### Tabulation and Summarize {#sec-tidyverse-tabulation}

Another useful but difficult task in base R is producing summary information by group. E.g., you want to produce a table that shows the average sepal length for each species. Here's how to do it using two new functions `group_by()` and `summarize()`, and pipe them together.

E.g., tally the number of samples by `Plant` and 'Type' in `CO2`. The function that does the tallying is `n()`.

```{r}
CO2 %>%
  group_by(Plant, Type) %>%
  summarize(n=n())
```

You do not see much difference between base R and `tidyverse` in the example above. But if you want to do more than tallying, `summarize()` is a treat, as `summarize()` supports the descriptive statistics functions mentioned before, such as mean(), sd(), median(), quantile(), etc. E.g,

```{r}
iris %>%
  group_by(Species) %>%
  summarize(u=mean(Sepal.Length))
```

You can apply multiple functions in `summarize()`. E.g., shows the number of samples and standard deviation.

```{r}
iris %>%
  group_by(Species) %>%
  summarize(n=n(),
            u=mean(Sepal.Length),
            s=sd(Sepal.Length))
```

You can group by more than one column. Let's switch to the `CO2` dataset to illustrate this point.

```{r}
CO2 %>%
  group_by(Plant, Type, Treatment) %>%
  summarize(n=n(),
            u=mean(conc),
            s=sd(conc))
```

**Two remarks:**

1.  `group_by()` is never used alone. It is often followed by `summarize()`.

2.  The type of variables fed to `group_by()` must be factors, i.e., categorical.

### A Summary of `tidyverse` functions

- `filter()`: select rows that satisfy the specified conditions.

- `select()`: choose or drop certain columns.

- `arrange()`: sort rows by the specified columns.

- `rename()`: change the name of a column.

- `mutate()`: add a column or modify the values of an existing column.

- `group_by()`, followed by `summarize()`

- pipe `%>%`: funnel intermediate result from an upstream process to the input of an immediate downstream process.

## Exercises for `tidyverse`

Q4. Repeat the exercises in Q1 using `tidyverse`.

Q5. Repeat the exercises in Q2 using `tidyverse`.

Q6. Repeat the exercises in Q3 using `tidyverse`.

Q7.

## Summary

In this chapter, you have seen how to perform data wrangling with basic R and `tidyverse`. What you have seen is only a fraction of what `tidyverse` can do. Readers who want to explore the full capability of `tidyverse` should consult the most authoritative source: [**R for Data Science**](https://r4ds.had.co.nz/) by Hadley Wickham & Garrett Grolemund. In the next chapter, you will see how to plot data. A picture is worth a thousand words. Data visualization is an indispensable step for data analysis. But before the next topic, let's recall all the new R commands discussed in this chapter.

```{=html}
<!--
`|`, `&`, `==`, `!`, `!=`, `subset()`, `sample()`, `order()`, `NULL` (a keyword), `table()`, `library()`, `require()`, `filter()`, `select()`, `arrange()`, `rename()`, `mutate()`, `group_by()`, `summarize()`, `n()`, and pip `%>%`.
-->
```

```{r echo=FALSE, warning=FALSE}
cmds <- c("|", "`&`", "`==`", "`!`", "`!=`", "`subset()`", "`sample()`", "`order()`", "`NULL` (a keyword)", "`table()`", "`library()`", "`require()`", "`filter()`", "`select()`", "`arrange()`", "`rename()`", "`mutate()`", "`group_by()`", "`summarize()`", "`n()`","pipe `%>%`", "", "", "") # fill up the matrix with empty strings

cmds_mat <- matrix(cmds, ncol = 4)
knitr::kable(
  cmds_mat, booktabs = TRUE,
  caption = "A Summary of R Operators and Functions"
)
```
