# Statistical Tests {#sec-04-statistical-tests}

R was inherently designed for statistical analysis. Therefore, it is justified to devote a chapter to statistical tests. Here, we will discuss a list of classic statistical tests with an emphasis on the Null Hypothesis Significance Test (NHST). We assume that you have a sound background in undergraduate-level statistics, i.e., you know the t statistic, $\chi^2$ statistic, confidence interval, p-value, correlation, etc. For readers who need to refresh basic statistical concepts, there are many excellent introductory statistics textbooks freely available on the Internet[^04_statistical_tests-1] [^04_statistical_tests-2] [^04_statistical_tests-3]. Interested readers are recommended to consult them for further detail.

[^04_statistical_tests-1]: Introductory Statistics for Life and Biomedical Sciences: <https://openintro.org/book/biostat/>

[^04_statistical_tests-2]: Introduction to Modern Statistics: <https://openintro.org/book/ims/>

[^04_statistical_tests-3]: Handbook of Biological Statistics: <https://www.biostathandbook.com/>

The goal of this chapter, as well as the whole book, emphasizes using R - the how. Additionally, the tests assume the availability of raw data, i.e., data sets contain individual samples. If only summary statistics (sample size, mean, standard deviation, etc.) are available, you have to use formulas to calculate the corresponding statistics, standard errors, critical values, and confidence intervals. These calculations are basically arithmetic operations and will not be discussed in this chapter.

## Learning Objectives: {.unnumbered}

- Test a quantitative variable in one, two, or more groups.

- Test for one or two categorical variables.

- Linear regression

## Housekeeping

We use two new data sets in this chapter: `FlightDelays`, and `Groceries` . Follow the steps below to import them into RStudio.

```{r}
if (!require(resampledata)) {
  install.packages('resampledata')
}
require(resampledata)
data("FlightDelays")
data("Groceries")

```

Take a glimpse of the data sets.

```{r}
summary(FlightDelays)
```

It contains 10 variables, out of which two are quantitative variables (`FlightLength`, and `Delay`), and the rest are categorical variables. Notice that there is a glitch: R mistakenly considered `FlightNo` as a quantitative variable. However, it does not affect us, as we do not use this variable here. If you want to convert it to a categorical variable (factor), here is the step:

```{r}
FlightDelays$FlightNo <- as.factor(FlightDelays$FlightNo)
```

The `Groceries` data set contains the product prices from Target and Walmart.

```{r}
summary(Groceries)
```

Lastly, we need to activate `tidyverse`.

```{r}
suppressMessages(require(tidyverse))
```

## Tests for Quantitative Variables

### One-sample t-tests {#sec-04-one-sample-t-test}

The goal of a one-sample t-test is to test if the mean of a quantitative variable is equal to a certain value. For example, if the aviation regulatory agency states that acceptable flight delay is 12 minutes. We want to see if the flights meet the requirement. The null hypothesis ($H_{0}$) is that the mean delay is equal to 12 minutes. The alternative hypothesis ($H_{A}$) is that the mean delay is not equal to 12 minutes. Here are the steps to perform such a test:

```{r}
t.test(FlightDelays$Delay, mu = 12)
```

As the p-value (0.6895) is greater than $\alpha$ = 0.05 at a 95% confidence level, we fail to reject the null hypothesis $H_{0}$, meaning that the mean flight delay meets the standard.

### One-sided t-tests

In the above flight delay example, the condition to reject the $H_{0}$ is that the delay time is either much longer or much shorter than 12 minutes (that is why the test is called two-sided). Indisputably, the latter case does not make much sense, as short delay times indicate the flights are punctual. Hence, the test must focus on long delay times by replacing the above two-sided test with a one-sided test in which the rejection region ($H_{A}$) lies at larger t-statistics (on the right). The parameter `alternative="greater"` suggests a one-sided test where the rejection region lies at the greater value.

```{r}
t.test(FlightDelays$Delay, mu = 12, alternative = "greater")
```

The p-value 0.6553 is larger than $\alpha$ = 0.05, fail to reject the $H_{0}$ at a 95% significance level.

### Two-sample t-tests {#sec-04-two-sample-t-test}

A two-sample t-test tests whether the mean of one group is the same as the other group. The $H_{0}$ states the two groups have the same mean, and the $H_{A}$ states otherwise.

For example, we want to compare the mean delays of the two carriers to determine whether they are the same.

*Solution 1:*

```{r}
AA_delay <- subset(FlightDelays, Carrier=="AA")
UA_delay <- subset(FlightDelays, Carrier=="UA")
t.test(AA_delay$Delay, UA_delay$Delay)
```

Based on the p-value (0.0001349) being less than $\alpha$ = 0.05, the delays of the two carriers are different. The bottom of the test results shows that the mean delay for AA (group x) is 10.09738, shorter than UA's 15.98308.

*Solution 2:*

This approach makes use of the categorical variable `carrier` to split the data set into two, avoiding the creation of two temporary data frames.

```{r}
t.test(Delay ~ Carrier, data=FlightDelays)
```

The symbol `~` means "split by". `Delay ~ Carrier` means to split `Delay` by `Carrier`.

### One-way ANOVA {#sec-04-one-way-ANOVA}

A one-way ANOVA tests whether the means of three or more groups are the same. Two key differences from the two-sample t-test above: 1. a two-sample t-test can handle only two groups, but ANOVA can handle two or more groups. 2. The $H_{0}$ of ANOVA is rejected if any pair(s) of groups have unequal means.

For example, test whether mean delays vary by destination.

```{r}
aov_results <- aov(Delay ~ Destination, data=FlightDelays)
summary(aov_results)
```

If you only care whether the $H_{0}$ can be rejected, focus on the overall p-value (2.29e-06) under the `Pr(>F)` column on the right. It is undeniably smaller than $\alpha$ = 0.05, so the $H_{0}$ can be rejected.

In some situations, you may be interested in finding out which pair(s) do not have the same mean. Tukey's Honest Significant Difference (HSD) test can answer this question.

```{r}
TukeyHSD(aov_results)
```

It lists all possible pairs of destinations. Since there are seven destinations, the total number of pairs is 21 (`choose(7,2)`).

Each row shows the difference of means (`diff`), the lower (`lwr`) and upper (`upr`) bounds of the confidence interval, and the adjusted p-value (`p adj`). Examining the p-value is the quickest way to identify pairs with unequal means. E.g., IAD-BNA in which the p-value is less than 0.05.

To ease analysis, it is suggested to plot the HSD result as below:

```{r}
hsd_results <- TukeyHSD(aov_results)

# las=1 causes the plot to print the pair name horizontally.
plot(hsd_results, las=1)
```

Confidence intervals not crossing zero (the vertical line) indicate pairs have unequal means.

### Paired t-tests {#sec-04-paired-t-test}

A paired t-test evaluates whether the mean difference between two measurements taken on the same subject significantly differs from zero. For a drug trial example, a subject's blood pressure is measured before and after taking the tested drug. Each sample is a subject. The measurements are blood pressure. The $H_{0}$ is that the mean difference between the two measurements is zero.

Here we will examine whether the product prices between Target and Walmart are different. The subject is a product. The two measurements are the prices of the same product from Target and Walmart.

```{r}
with(Groceries, t.test(Target, Walmart, paired=T))
```

The p-value of 0.6415 \> $\alpha$ = 0.05 suggests that there is no statistical difference between the two stores at the 95% confidence level. The mean price difference of all the products in the sample is 0.057, which is not considered a significant difference from zero.

## Tests for Categorical Variables

### One Categorical Variable {#sec-04-x2-gof-test}

Here, we want to check whether the proportions of categories match a specified expected ratio. In standard textbooks, this test is called the **Chi-square (**$\chi^{2}$**) goodness-of-fit** test. For example, `Delayed30` indicates whether the flight was delayed by more than 30 minutes; it is either Yes or No. Suppose the regulatory standard recommends not more than 10% of a random sample of flights. First, figure out how many Yes and No in the sample.

```{r}
table(FlightDelays$Delayed30)
```

The data set contains 4,029 flights; 3,432 were punctual, and 587 were delayed. According to the standard, the number of delayed flights should be no more than 10% of the total, or 402. Due to random variation in sampling, we cannot conclude an issue is present simply because 587 is greater than 402. Here's where the $\chi^2$-test comes in.

```{r}
chisq.test(table(FlightDelays$Delayed30), p=c(0.9,0.1))
```

The $H_{0}$ assumes the proportions agree with the expected ratio. The p-value shown above is literally zero, indicating that the proportions of Yes and No do not conform to the expected 9:1 ratio at the 95% confidence level.

One remark about setting up the proportion parameter `p=c(0.9,0.1)`. Make sure that the order of the proportions matches the sequence of the categories produced by the `table()` function. Additionally, the proportions must sum to 1.

### Two Categorical Variables {#sec-04-x2-test-independence}

Here we examine the independence between two categorical variables. The test is colloquially called $\chi^{2}$ **Test for Independence**. For example, we want to know whether a delay of more than 30 minutes depends on the carrier, that is to say whether `Delayed30` is independent of `Carrier`. The $H_{0}$ assumes that they are independent.

```{r}
with(FlightDelays, chisq.test(Carrier, Delayed30, correct=TRUE))
```

Note that `correct=TRUE` instructs the test function to make adjustments in calculating the $\chi^2$ statistic. The correction is only applicable for a 2x2 contingency table with small counts. However, it does little harm if you turn it on for tables with a different dimension or large counts[^04_statistical_tests-4]. My suggestion is to turn it on all the time.

[^04_statistical_tests-4]: <https://en.wikipedia.org/wiki/Yates%27s_correction_for_continuity>

As the p-value of 0.0002434 is less than 0.05, reject the $H_{0}$ at the 95% confidence level, meaning that `Delay30` and `Carrier` are not independent. This conclusion suggests that the proportion of delayed flights differs between the two carriers. To be clear, their dependency says nothing about the carriers' delay issue. To figure that out exactly, we should examine the proportion of delayed flights for each carrier using a new function `prop.table()` as below:

```{r}
with(FlightDelays, prop.table(table(Carrier, Delayed30), margin=1))
```

We feed the contingency table of `Carrier` by `Delayed30` to the `prop.table()` function with `margin=1`, meaning that values are divided by the row totals or the sum of the proportions for each row is 1. As you can see, both carriers have an issue under the 10% rule, but the issue with UA is more severe than AA.

## Regression

### Simple Linear Regression {#sec-04-SLR}

A simple linear regression (SLR) model involves a predictor X (independent variable) and a response Y (dependent variable), where both of them are quantitative, but they have different roles.

$$Y = \beta_{0} + \beta_{1}X$$

where $\beta_{0}$ is the intercept, and $\beta_{1}$ is the slope.

The relationship between them in an SLR is linear and directional. Firstly, the predictor is believed to influence the response, and not the reverse; therefore, it is directional. For example, temperature (predictor) is supposed to influence ice cream sales (response), and not the other way.

Secondly, the relationship is linear, meaning that a unit increment of X will cause Y to increase or decrease by $\beta_{1}$ units for every X within the range of X in the data.

Thirdly, an SLR can only establish a correlation between X and Y, but not causation.

Finally, in the context of NHST, the $H_{0}$ is $\beta_{1}$ = 0.

Here we will switch to the familiar built-in `iris` data set to demonstrate how to build an SLR. Suppose we are interested in whether `Sep.Length` (predictor) affects `Sep.Width` (response) for `setosa` species. The first step for building an SLR is to visualize the relationship between the two variables using a scatter plot (@sec-baseR-scatter-plot or @sec-ggplot-scatter-plot). Do not downplay the intuition of visual inspection. In the absence of an upward or downward trend, the attempt to build an SLR is deemed fruitless. On a positive note, the plot may still reveal an intriguing trend not as expected, requiring a different kind of model or variable transformation. Anyway, let's plot the two variables.

```{r}
setosa_iris <- subset(iris, Species=='setosa')
ggplot(data=setosa_iris, aes(x=Sepal.Length, y=Sepal.Width)) +
  geom_point()
```

As shown above, there is an upward trend even though it is not definitive for everyone. Subjectively, we will proceed to build an SLR model.

```{r}
slr_model <- lm(Sepal.Width ~ Sepal.Length, data=setosa_iris)
summary(slr_model)
```

The slope and the intercept can be found under the `Estimate` column of `Coefficients` subsection. In this example, the intercept $\beta_{0}$ is -0.5694, and the slope $\beta_{1}$ (`Sepal.Length`) is 0.7985. So the SLR model is:

$$Sepal.Width = -0.5694 + 0.7985 \times Sepal.Length   $$

The SLR model says that a unit increase in `Sepal.Length`, `Sepal.Width` will increase by 0.7985 units.

To judge whether they are statistically significant, examine the p-values under the column `Pr(>|t|)`. The p-value of the slope is close to zero, suggesting that it is statistically significant; unfortunately, the intercept is not. It means that the linear correlation between `Sepal.Width` and `Sepal.Length` is strong. However, it is completely accurate to predict `Sepal.Width` using `Sepal.Length`.

Since we mentioned correlation, the `Multiple R-squared` is the square of Pearson's correlation between the two variables. Let's verify this.

```{r}
# Pearson's correlation
r <- with(setosa_iris, cor(Sepal.Length, Sepal.Width))

print(paste("Pearson's correlation =",r))

# Square of Pearson's correlation is Multiple R-squared
print(paste("The square of r =",r**2))
```

### Multiple Linear Regression {#sec-04-MLR}

A multiple linear regression (MLR) model involves one response and two or more predictors. The equation for MLR is

$$Y = \beta_{0} + \beta_{1} \times X_{1} + ... + \beta_{n} \times X_{n}$$

Suppose we are keen to predict `Sepal.Width` by `Sepal.Length` and `Petal.Length` for setosa iris. Similar to SLR, we visualize the data but using a multipanel scatter plot @sec-baseR-multipanel-scatter-plot, @sec-ggplot-multipanel-scatter-plot).

```{r}
pairs(setosa_iris[,1:4], main="A Multi-Panel Scatter Plot")
```

Here we visualize the trend between `Sepal.Width` and `Sepal.Length` (2^nd^ row and 1^st^ column), `Sepal.Width` and `Petal.Length` (2^nd^ row and 3^rd^ column), and `Sepal.Width` and `Petal.Width` (2^nd^ row and 4^th^ column). Despite the trend not being clear between `Sepal.Width` and `Petal.Length`, and `Sepal.Width` and `Petal.Width`, let's proceed to build an MLR model.

```{r}
mlr_model <- lm(Sepal.Width ~ Sepal.Length + Petal.Length + Petal.Width, data=setosa_iris)
summary(mlr_model)
```

Based on the p-values, only `Sepal.Length` is a stable predictor of `Sepal.Width`; `Petal.Length` and `Petal.Width` do not make a statistically significant contribution to the model. This example is simple, as there are not many predictors to be considered. When the number of predictors rises, the number of choices explodes exponentially. For a four-predictor example, the number of combinations is:

$$\sum_{i=1}^{n=4} \binom{n}{i}\ = \binom{4}{1}\ + \binom{4}{2}\ + \binom{4}{3}\ + \binom{4}{4}\ = 1 + 6 + 6 + 1 = 14$$ R has a function to assist in finding a "not too bad" MLR model, meaning it does not guarantee to return the best model. Here is how it works. It begins with a one-predictor model; start with any predictor. Then it tries to build an intermediate model by incorporating an additional predictor from the pool that returns the least error. This step will continue until all predictors are chosen.

```{r}
baseline_mlr <- lm(Sepal.Width ~ Sepal.Length, data=setosa_iris)
add1(baseline_mlr, scope = ~ Sepal.Length + Petal.Length + Petal.Width, test = "F")
```

The model error is reflected in the Akaike Information Criterion (AIC) values. The intuition is that more predictors will naturally fit the data better. However, the model inevitably fits the noise as well. As a result, although the model looks great in development, it eventually fails to handle unseen data, i.e., overfitting. AIC serves as a canary. As a rule of thumb, if the difference in AIC between consecutive intermediate models is less than 2, there is no significant difference.

## Summary

In this chapter, we have discussed classical statistical tests in the framework of Null Hypothesis Significance Test (NHST). A simple guideline for choosing the most appropriate test is to identify the main variable(s), how many there are, and their type(s). The table below summarizes the tests discussed in this chapter:

| Name | Description | Example | Reference |
|------------------|------------------|------------------|------------------|
| One-sample t-test | Test whether the mean of a quantitative variable equals a specified value. | `t.test(FlightDelays$Delay, mu = 12)` | @sec-04-one-sample-t-test |
| Two-sample t-test | Test whether or not the means of two groups are equal. | `t.test(Delay ~ Carrier, data=FlightDelays)` | @sec-04-two-sample-t-test |
| One-way ANOVA | Test whether or not the means of three or more groups are equal. | `aov_results <- aov(Delay ~ Destination, data=FlightDelays)` | @sec-04-one-way-ANOVA |
| Paired t-test | Test whether the mean difference between two measurements taken on the same subject significantly differs from zero | `with(Groceries, t.test(Target, Walmart, paired=T))` | @sec-04-paired-t-test |
| $\chi^2$ goodness-of-fit | Test whether the proportions of categories match a specified expected ratio. | `chisq.test(table(FlightDelays$Delayed30), p=c(0.9,0.1))` | @sec-04-x2-gof-test |
| $\chi^2$ test for independence | Test the independence between two categorical variables | `with(FlightDelays, chisq.test(Carrier, Delayed30, correct=TRUE))` | @sec-04-x2-test-independence |
| Simple Linear Regression | It examines the linear relationship between the predictor and the response. | `lm(Sepal.Width ~ Sepal.Length, data=setosa_iris)` | @sec-04-SLR |
| Multiple Linear Regression | It examines the linear relationship between multiple predictors and the response. | `lm(Sepal.Width ~ Sepal.Length + Petal.Length, data=setosa_iris)` | @sec-04-MLR |

## Exercises

Use the built-in data set `mtcars` to answer these questions. Before tackling the questions below, you need to convert the categorical variables in `mtcars` to a factor. Here is the code:

```{r}
mtcars$cyl <- factor(mtcars$cyl)
mtcars$vs <- factor(mtcars$vs, levels = c(0,1))
mtcars$am <- factor(mtcars$am, levels = c(0,1))
```

Q1. Test whether the mean `mpg` is 20 at the 95% confidence level.

```{r echo=F, eval=F}
t.test(mtcars$mpg, mu = 20)
# p-value = 0.93, fail to reject the null at 95%
```

Q2. `am` stands for automatic transmission. A 0 means not automatic and 1 means automatic. Test whether the mean acceleration `qsec` is the same between automatic and manual transmissions.

```{r echo=F, eval=F}
t.test(qsec ~ am, data=mtcars)

# p-value=0.2093, fail to reject the null at 95% C.L., meaning that the acceleration is the same between the two types of transmission.
```

Q3. Test whether the mean acceleration `qsec` is the same for different numbers of cylinders `cyl`. If it is not true, identify the pair(s) that have different mean acceleration. Note that `cyl` must be converted to a factor before performing any tests.

```{r echo=F, eval=F}
qsec_aov <- aov(qsec ~ cyl, data=mtcars)
summary(qsec_aov)

# the p-value is 0.00196<0.05, reject the null at 95% C.L.
```

```{r echo=F, eval=F}
qsec_hsd <- TukeyHSD(qsec_aov)

plot(qsec_hsd, las=1)
# 8 cylinders vs 4 cylinders is the pair that is tested to have unequal mean acceleration, as it did cross the line = 0. 
```

Q4. Test whether cars with different numbers of cylinders are equally sampled in the data set.

```{r echo=F, eval=F}
chisq.test(table(mtcars$cyl), p = c(1/3,1/3,1/3))

# the p-value is 0.3147 < 0.05, fail to reject the null at 95% C.L. It means that the numbers of cars for each cylinder are not exactly equal, but they are probably due to random sampling.
```

Q5. Test whether the type of transmission `am` and the number of cylinders are independent. Note: If you encounter a warning message, it is due to small-count data. You can ignore it.

```{r echo=F, eval=F}
chisq.test(with(mtcars, table(cyl, am)), correct=T)

# p-value = 0.01265 < 0.05, reject the null at 95% C.L., meaning that the mode of transmission and the number of cylinders are not independent.
```

Q6. Use a linear model to test the relationship between horsepower `hp` and acceleration `qsec`.

```{r echo=F, eval=F}
# hp: predictor
# qsec : response

# Take a look at the data
plot(x=mtcars$hp, y=mtcars$qsec, xlab="Horsepower", ylab="Acceleration in seconds")

# There is an inverse relationship.
```

```{r echo=F, eval=F}
hp_qsec_lm <- lm( qsec ~ hp, data=mtcars)
summary(hp_qsec_lm)

# small pvalues for both the intercept and slope.
# cor = sqrt(0.5) ~ -0.7 bc the slope is negative.
```

```{r echo=F, eval=F}

plot(x=mtcars$hp, y=mtcars$qsec, xlab="Horsepower", ylab="Acceleration in seconds")
intercept <- hp_qsec_lm$coefficients[1]
slope <- hp_qsec_lm$coefficients[2]
abline(a=intercept, b=slope, col='red')
```

Q7. Use a multiple linear model to study the relationship between acceleration, horsepower, weight, and MPG.

```{r echo=F, eval=F}

# Take a peek of the data
pairs(mtcars[,c('qsec','hp','wt','mpg')])

# it seems qsec is linear correlated with hp and mpg, but not much with wt.
```

```{r echo=F, eval=F}

mlr_model <- lm(qsec ~ hp + wt + mpg, data=mtcars)
summary(mlr_model)

# p-values of hp < wt < mpg. So, hp is the best predictor, followed by wt, and mpg.
```
